\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks

### 2015-10-19 nnt_manu.Rnw
Sehan edits

%% an abstract and keywords
\Abstract{
Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.  A common criticism of the method is the inability to characterize relationships among variables from a fitted model. Although several methods proposed in recent years to `illuminate the black-box', they have not been made available in an open-source programming environment.  This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.  Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.  Methods are provided for objects from many of the common neural networks packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.  Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.      
}
\Keywords{neural networks, plotnet, sensitivity, variable importance, \proglang{R}}
\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
%% at least one keyword must be supplied

@@

Training a neural network model requires identifying the `optimal' weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in \citep{Rumelhart86}.  This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.  Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Ripley96}:
\begin{equation}
y_k = f_k \left(\sum\limits_{j=1}^k w_{jk}f_j \left( \sum\limits_{i=1}^j w_{ij}x_i\right) \right)
\end{equation}
where the estimated value of the response variable $y_k$ is a summation of the product between multiple input variables $x$ and multiple hidden nodes $j$, as mediated by the weights $w$ and activation functions $f_j$ and $f_k$ for each hidden and output node.

@@

A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection \citep{Ripley96}. Models with skip layers include additional connections from the input to output layers that bypass the hidden layer.  The default behavior of \code{plotnet} is to plot the primary network, whereas the skip layer can be viewed separately with \code{skip = TRUE}. If \code{nid = TRUE}, the line widths for both the primary and skip layer plots are relative to all weights. Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0} in \code{nnet}) will include bias connections to the output layer, whereas these are included only in the primary plot if size is greater than zero.

@@

\begin{figure}
\centering
\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
<<plotimp-1, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
garson(mod1)
@
\label{fig:plotimp1}
}
\subfloat[][Model 1 (\code{mlp}), \code{olden}]{
<<plotimp-2, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
olden(mod1)
@
\label{fig:plotimp2}
}

@@

Although \pkg{NeuralnetTools} provides several methods to extract information from a fitted neural network, it does not provide explicit guidance for developing the initial model.  A potentially more challenging aspect of using \ac{mlp} neural networks is understanding the effects of network architecture on model performance, appropriate use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting \citep{Maier00}.  A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to emphasize the importance of these considerations.  The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive or practical application of model development.  In general, the following should be considered during initial development \cite{Ripley96, Lek00, Maier00}:
\begin{itemize}
\item Initial data pre-processing to normalize inputs, standardize response, or assess influence of outliers,
\item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, pruning weights or inputs,
\item Separating data into training and test datasets, e.g., 2:1, 3:1, leave-one-out, etc., 
\item Initial starting weights for the back-propagation algorithm, and
\item Criteria for stopping model training, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
\end{itemize}

@@

\begin{figure}
\centering
\subfloat[][Networks with one node]{
<<flightimp-1, fig.height = 3, fig.width = 8, out.width = '0.95\\textwidth', echo = F>>=
print(p1)
@
\label{fig:flightimp1}
}

@@

I thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the University of Minnesota for general guidance during the development of this package.  Contributions and suggestions from online users have also greatly improved the utility of the package.  Funding for this project was supported in part by an Interdisciplinary Doctoral Fellowship provided by the Graduate School at the University of Minnesota to M. Beck.

>>>>

%% an abstract and keywords
\Abstract{
Supervised neural networks have been applied as a machine learning technique to identify and predict emergent patterns in unstructured datasets.  A common criticism of the method is the inability to characterize relationships among variables from a fitted model. Although several methods have been proposed to `illuminate the black-box', they have not been made available in an open-source programming environment.  This article describes the \pkg{NeuralNetTools} package that can be used for the interpretation of supervised neural network models created in \proglang{R}.  Functions in the package can be used to visualize a model using a neural network interpretation diagram, evaluate variable importance by disaggregating the model weights, and perform a sensitivity analysis of the response variables to changes in the input variables.  Methods are provided for objects from many of the common neural networks packages in \proglang{R}, including \pkg{caret}, \pkg{neuralnet}, \pkg{nnet}, and \pkg{RSNNS}.  The article provides a brief overview of the theoretical foundation of neural networks, a description of the package structure and functions, and an applied example to provide a context for model development with \pkg{NeuralNetTools}.  Overall, the package provides a toolset for neural networks that complements existing quantitative techniques for data-intensive exploration.      
}
\Keywords{neural networks, plotnet, sensitivity, variable importance, \proglang{R}}
\Plainkeywords{neural networks, plotnet, sensitivity, variable importance, R} %% without formatting
%% at least one keyword must be supplied

@@

Training a neural network model requires identifying the `optimal' weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in \citep{Rumelhart86}.  This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.  Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
\begin{equation}
y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right)
\end{equation}
where the estimated value $k$ of the response variable $y$ is a summation of the products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.

@@

A primary and skip layer network can also be plotted for \code{nnet} models with a skip layer connection (\cref{fig:plotnet_skip}). Models with skip layers include additional connections from the input to output layers that bypass the hidden layer \citep{Ripley96}.  The default behavior of \code{plotnet} is to plot the primary network, whereas the skip layer can be viewed separately with \code{skip = TRUE}. If \code{nid = TRUE}, the line widths for both the primary and skip layer plots are relative to all weights. Plotting a network with only a skip layer (i.e., no hidden layer, \code{size = 0} in \code{nnet}) will include bias connections to the output layer, whereas these are included only in the primary plot if size is greater than zero.

@@

\begin{figure}[!h]
\centering
\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
<<plotimp-1, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
garson(mod1)
@
\label{fig:plotimp1}
}
\subfloat[][Model 1 (\code{mlp}), \code{olden}]{
<<plotimp-2, fig.height = 2, fig.width = 4, out.width = '0.47\\textwidth', echo = F>>=
olden(mod1)
@
\label{fig:plotimp2}
}

@@

Although \pkg{NeuralnetTools} provides several methods to extract information from a fitted neural network, it does not provide explicit guidance for developing the initial model.  A potentially more challenging aspect of using \ac{mlp} neural networks is understanding the effects of network architecture on model performance, appropriate use of training and validation datasets, and implications for the bias-variance tradeoff with model over- or under-fitting \citep{Maier00}.  A detailed discussion of these issues is beyond the scope of this paper, although an example application is presented below to emphasize the importance of these considerations.  The models presented above, including the \code{neuraldat} dataset, are contrived examples to illustrate use of the \pkg{NeuralNetTools} package and they do not demonstrate a comprehensive or practical application of model development.  In general, the following should be considered during initial development \cite{Ripley96, Lek00, Maier00}:
\begin{itemize}
\item Initial data pre-processing to normalize inputs, standardize response, and assess influence of outliers,
\item Network architecture including number of hidden layers, number of nodes in each hidden layer, inclusion of bias or skip layers, and pruning weights or inputs,
\item Separating data into training and test datasets, e.g., 2:1, 3:1, leave-one-out, etc., 
\item Initial starting weights for the back-propagation algorithm, and
\item Criteria for stopping model training, e.g., error convergance tolerance, maximum number of iterations, minimum error on test dataset, etc.
\end{itemize}

@@

\begin{figure}[!h]
\centering
\subfloat[][Networks with one node]{
<<flightimp-1, fig.height = 3, fig.width = 8, out.width = '0.95\\textwidth', echo = F>>=
print(p1)
@
\label{fig:flightimp1}
}

@@

I thank Bruce Vondracek, Sanford Weisberg, and Bruce Wilson of the University of Minnesota for general guidance during the development of this package.  Thanks to Sehan Lee for reviewing an earlier draft.  Contributions and suggestions from online users have also greatly improved the utility of the package.  Funding for this project was supported in part by an Interdisciplinary Doctoral Fellowship provided by the Graduate School at the University of Minnesota to M. Beck.

### 2015-10-19 nnt_manu.tex
Sehan edits

Training a neural network model requires identifying the `optimal' weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in \citep{Rumelhart86}.  This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.  Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Ripley96}:
\begin{equation}
y_k = f_k \left(\sum\limits_{j=1}^k w_{jk}f_j \left( \sum\limits_{i=1}^j w_{ij}x_i\right) \right)
\end{equation}
where the estimated value of the response variable $y_k$ is a summation of the product between multiple input variables $x$ and multiple hidden nodes $j$, as mediated by the weights $w$ and activation functions $f_j$ and $f_k$ for each hidden and output node.

Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 78\%.  As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.  Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class.  Additional \code{default} or \code{numeric} methods are available for some of the generic functions.

@@

\begin{figure}
\centering
\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}

@@

\begin{figure}
\centering
\subfloat[][Networks with one node]{
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}

>>>>

Training a neural network model requires identifying the `optimal' weights that define the connections between the model layers.  The optimal weights are those that minimize prediction error for a test dataset that is independent of the training dataset.  Training is commonly achieved using the backpropagation algorithm described in \citep{Rumelhart86}.  This algorithm identifies the optimal weighting scheme through an iterative process where weights are gradually changed through a forward- and backward-propagation process \citep{Rumelhart86,Lek00}.  The algorithm begins by assigning an arbitrary weighting scheme to the connections in the network, followed by estimating the output in the response variable through the forward-propagation of information through the network, and finally calculating the difference between the predicted and actual value of the response.  The weights are then changed through a back-propagation step that begins by changing weights in the output layer and then the remaining hidden layers.  The process is repeated until the chosen error function is minimized, as in standard model-fitting techniques for regression \citep{Cheng94}.  Formulaically, a fitted \ac{mlp} neural network can be represented as \citep{Bishop95,Venables02}:
\begin{equation}
y_k = f_o \left(\sum\limits_{h} w_{hk}f_h \left( \sum\limits_{i} w_{ih}x_i\right) \right)
\end{equation}
where the estimated value $k$ of the response variable $y$ is a summation of the products between the respective weights $w$ for $i$ input variables $x$ and $h$ hidden nodes, mediated by the activation functions $f_h$ and $f_o$ for each hidden and output node.

Methods in \pkg{NeuralNetTools} were written for several \proglang{R} packages that can be used to create \ac{mlp} neural networks: \pkg{neuralnet} \citep{Fritsch12}, \pkg{nnet} \citep{Venables02}, and \pkg{RSNNS} \citep{Bergmeir12}. Limited methods were also developed for neural network objects created with the \code{train} function from the \pkg{caret} package \citep{Kuhn15}.  Additional \proglang{R} packages that can create \ac{mlp} neural networks include \pkg{AMORE} that implements the ``TAO-robust back-propagation algorithm'' for model fitting \citep{Castejon14}, \pkg{FCNN4R} as an \proglang{R} interface to the \pkg{FCNN} \proglang{C++} library \citep{Klima15}, \pkg{monmlp} for networks with partial monotonicity constraints \citep{Cannon15}, and \pkg{qrnn} for quantile regression neural networks \citep{Cannon11}.  At the time of writing, the \ac{CRAN} download logs \citep{Csardi15} showed that the \proglang{R} packages with methods in \pkg{NeuralNetTools} included 94\% of all downloads for the available \ac{mlp} packages, with \pkg{nnet} accounting for over 79\%.  As such, methods have not been included in \pkg{NeuralNetTools} for the remaining packages, although further development of \pkg{NeuralNetTools} could include additional methods based on popularity.  Methods for each function are currently available for  \code{mlp} (\pkg{RSNNS}), \code{nn} (\pkg{neuralnet}), \code{nnet} (\pkg{nnet}), and \code{train} (\pkg{caret}, only if the object also inherits from the \code{nnet} class.  Additional \code{default} or \code{numeric} methods are available for some of the generic functions.

@@

\begin{figure}[!h]
\centering
\subfloat[][Model 1 (\code{mlp}), \code{garson}]{
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}

@@

\begin{figure}[!h]
\centering
\subfloat[][Networks with one node]{
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}
